---
title: "AP_ASSIGNMENT_20415440"
author: "Zayd Orabi"
date: "`r Sys.Date()`"
output: word_document
---


# PART 1

Write the code to analyse the distribution of COVID patients (confirmed
or suspected) across counties. Write the code to investigate the
distribution of the patients across age groups (e.g., 0-18, 19-35,
36-50, 51+). Visualise both the findings using the histogram. Explain
your findings.

```{r}
#Loading libraries
library(DBI)
library(dplyr)
library(ggplot2)
```

```{r}
#Loading datasets
patients <- read.csv("patientsUG.csv")
encounters <- read.csv("encountersUG.csv")
conditions <- read.csv("conditionsUG.csv")
```

```{r}
head(patients)
head(encounters)
head(conditions)
```

```{r}
#Combining data sets to extract necessary data
part1_covid <- patients %>%
  inner_join(encounters, by = c("Id" = "PATIENT")) %>%
  filter(REASONDESCRIPTION %in% c("COVID-19", "Suspected COVID-19")) %>%
  distinct(Id, .keep_all = TRUE)

head(part1_covid)
```

```{r}
#Analyze the distribution across counties
county_distribution <- part1_covid %>%
  group_by(COUNTY) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

print(county_distribution)
```

```{r}
#Visualising the distribution.
ggplot(county_distribution, aes(x = reorder(COUNTY, -count), y = count)) +
  geom_bar(stat = "identity") +
  coord_flip() + #Flipped the Axis so county labels dont overlap
  labs(title = "Distribution of COVID-19 Cases Across Counties",
       x = "County",
       y = "Number of Cases")
```

The bar plot shows an uneven distribution of COVID-19 cases across
counties, with some counties being more affected than others. Middlesex
County stands out as an outlier with almost double the amount of cases
of any other county. The differences in the amount of cases could be due
to factors like population density or measures taken within that county
to prevent the spread of covid.

```{r}
#Calculate age of each patient
part1_covid_age <- part1_covid %>%
  mutate(BIRTHDATE = as.Date(BIRTHDATE),
         AGE = as.integer(floor((Sys.Date() - BIRTHDATE) / 365)))

head(part1_covid_age)
```

```{r}
#Categorize into age groups
part1_covid_agegroup <- part1_covid_age %>%
  mutate(AGE_GROUP = case_when(
    AGE <= 18 ~ "0-18",
    AGE <= 35 ~ "19-35",
    AGE <= 50 ~ "36-50",
    AGE > 50 ~ "51+"
  ))

head(part1_covid_agegroup)
```

```{r}
#Display the distribution across age groups
agegroup_distribution <- part1_covid_agegroup %>%
  group_by(AGE_GROUP) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

print(agegroup_distribution)
```

```{r}
#Visualising the distribution
ggplot(part1_covid_agegroup, aes(x = AGE_GROUP)) +
  geom_histogram(stat = "count") +
  labs(title = "Distribution of COVID-19 Patients Across Age Groups",
       x = "Age Group",
       y = "Number of Patients") +
  theme_minimal()
```

The distribution of patients across age groups is not normal. The
significantly higher number of admitted patients in the 51+ age group
highlights the vulnerability of older adults to COVID-19. The high
number of cases in the 19-35 age group may result from greater social
interaction, work-related exposures, and possibly lower adherence to
preventive measures.

# PART 2

Filter those patients in the dataset that have contracted COVID-19 or
Suspected COVID-19; ; what are the top 10 most common conditions
(symptoms) related to the patients? Do the conditions differ between
genders? Provide a table to rank the top 10 conditions for male and
female patients separately. Elaborate on the findings.

```{r}
#Filter conditions related to COVID-19 or Suspected COVID-19
covid_conditions <- conditions %>%
  filter(DESCRIPTION %in% c("COVID-19", "Suspected COVID-19"))

head(covid_conditions)
```

```{r}
#Filter the conditions dataset for the unique patient Ids
covid_patient_ids <- unique(covid_conditions$PATIENT)

covid_related_conditions <- conditions %>%
  filter(PATIENT %in% covid_patient_ids & ENCOUNTER %in% covid_conditions$ENCOUNTER)

head(covid_related_conditions)
```

```{r}
#Join with patients data set for gender
part2_conditions <- merge(covid_related_conditions, patients, by.x = "PATIENT", by.y = "Id")

head(part2_conditions)
```

```{r}
#Count the conditions
count_conditions <- part2_conditions %>%
  group_by(DESCRIPTION) %>%
  filter(!DESCRIPTION %in% c("COVID-19", "Suspected COVID-19")) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

print(count_conditions)
```

the top 10 most common COVID-19 related symptoms are: 
1. Fever 
2. Cough
3. Loss of taste 
4. Fatigue 
5. Sputum finding 
6. Dyspnea 
7. Wheezing 
8. Joint Pain 
9. Muscle Pain 
10. Sore throat

```{r}
#Count the conditions for male patients
count_conditions_male <- part2_conditions %>%
  filter(GENDER == "M") %>%
  filter(!DESCRIPTION %in% c("COVID-19", "Suspected COVID-19")) %>%
  group_by(DESCRIPTION) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

print(count_conditions_male)
```

the top 10 most common COVID-19 related symptoms for males are: 1. Fever
2. Cough 3. Loss of taste 4. Fatigue 5. Sputum finding 6. Dyspnea 7.
Wheezing 8. Joint Pain 9. Muscle Pain 10. Headache

```{r}
#Count the conditions for female patients
count_conditions_female <- part2_conditions %>%
  filter(GENDER == "F") %>%
  filter(!DESCRIPTION %in% c("COVID-19", "Suspected COVID-19")) %>%
  group_by(DESCRIPTION) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

print(count_conditions_female)
```

the top 10 most common COVID-19 related symptoms for females are: 1.
Fever 2. Cough 3. Loss of taste 4. Fatigue 5. Sputum finding 6. Dyspnea
7. Wheezing 8. Joint Pain 9. Muscle Pain 10. Sore throat

The similar top symptoms for both genders indicate that COVID-19
symptoms are common across the board. However, specific differences,
such as sore throat being more common in females and headache in males,
suggest variations in disease manifestation or reporting, possibly due
to biological or social factors.

# PART 3

Write the code to analyse the factors that might influence the
hospitalisation rate for the COVID patient (confirmed or suspected) in
the dataset. Any factors in the dataset, such as age, gender, zip code,
marital status, race and county, can be considered. Pick 2 of the
factors and explain if there is a trend that explains the variation.

```{r}
part3_covid <- patients %>%
  inner_join(encounters, by = c("Id" = "PATIENT")) %>%
  filter(REASONDESCRIPTION %in% c("COVID-19")) 

head(part3_covid)
```

```{r}
#Separating county and encounter class for bar graph
hospitalization1 <- part3_covid %>%
  group_by(COUNTY, ENCOUNTERCLASS) %>%
  summarise(count = n()) %>%
  mutate(rate = count / sum(count) * 100)

print(hospitalization1)
```

```{r}
#Plotting bar graph
ggplot(hospitalization1, aes(x = COUNTY, y = rate, fill = ENCOUNTERCLASS)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Hospitalization Rate by County", x = "County", y = "Rate (%)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The dominance of inpatient encounters across counties suggests that
COVID-19 cases requiring significant medical intervention are
widespread, affecting all regions similarly. The variation in wellness
encounters might be influenced by factors such as healthcare access,
population demographics, and regional healthcare policies. Counties with
higher wellness encounter rates may have better access to healthcare
facilities, leading to more routine check-ups and non-emergency visits.

```{r}
#Separating county and encounter class for bar graph
hospitalization2 <- part3_covid %>%
  group_by(GENDER, ENCOUNTERCLASS) %>%
  summarise(count = n()) %>%
  mutate(rate = count / sum(count) * 100)

head(hospitalization2)
```

```{r}
#Plotting bar graph
ggplot(hospitalization2, aes(x = GENDER, y = rate, fill = ENCOUNTERCLASS)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Hospitalization Rate by Gender", x = "Gender", y = "Rate (%)")
```

The higher rate of inpatient encounters suggests that severe cases
requiring hospital admission are more common among both genders, but
slightly more prevalent in females. The higher wellness encounter rate
in males could indicate a greater frequency of routine check-ups or
non-emergency visits among male patients.

# PART 4

```{r}
part2_conditions <- part2_conditions %>%
  mutate(
    START = as.Date(START, format="%Y-%m-%d"),
    STOP = as.Date(STOP, format="%Y-%m-%d"),
    BIRTHDATE = as.Date(BIRTHDATE, format="%Y-%m-%d"),
    DEATHDATE = as.Date(DEATHDATE, format="%Y-%m-%d", errors = "coerce")
  )
```


```{r}
#Calculate age at the time of diagnosis
part4_conditions <- part2_conditions %>%
  mutate(
    AGE_AT_DIAGNOSIS = as.numeric(difftime(START, BIRTHDATE, units = "days")) / 365
  )

head(part4_conditions)
```

```{r}
#Add a column to determine if the patient recovered
part4_conditions <- part4_conditions %>%
  mutate(
    RECOVERED = ifelse(!is.na(STOP), "Recovered", "Not Recovered")
  )

head(part4_conditions)
```

```{r}
#Analyze demographics, symptoms, and timeline of diagnosis and recovery
part4_summary <- part4_conditions %>%
  distinct(PATIENT, .keep_all = TRUE) %>%
  group_by(RECOVERED) %>%
  summarise(
    avg_age = mean(AGE_AT_DIAGNOSIS, na.rm = TRUE),
    gender_distribution = list(table(GENDER)),
    race_distribution = list(table(RACE)),
    avg_recovery_duration = mean(as.numeric(difftime(STOP, START, units = "days")), na.rm = TRUE)
  )

print(part4_summary)
```

```{r}
print(part4_summary$gender_distribution)
```

```{r}
print(part4_summary$race_distribution)
```


```{r}
#Additional analysis on the symptoms
part4_symptoms <- part4_conditions %>%
  group_by(RECOVERED, DESCRIPTION) %>%
  summarise(
    symptom_count = n()
  ) %>%
  arrange(RECOVERED, desc(symptom_count))

print(part4_symptoms)
```

Based on the data analysis, several key observations can be made. Firstly, younger individuals are more likely to recover from COVID-19, likely due to stronger immune systems and fewer preexisting conditions. Secondly, females have a higher recovery rate than males. This might be due to biological differences, social behaviors, and possibly different levels of exposure to the virus. Females may also engage in more health-conscious behaviors, contributing to their higher recovery rates. Thirdly, Native Americans appear to have the highest recovery rate, though the sample size is too small for definitive conclusions. Among larger racial groups, Asians have the highest recovery rate at 94%, while African Americans have the lowest at 92%. These disparities could be influenced by factors such as socioeconomic status, access to healthcare, and living conditions. Additionally, social determinants of health, including employment types, living environments, and education levels, may also play significant roles in these differences.

# END











































